---
title: "Unit2"
author: "K Iwasaki"
date: "September 13, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Example

```{r}
df = read.table(file = "Placekick.csv", header = TRUE, sep = ",")
str(df)
head(df)

# check NA in each column
apply(is.na(df), 2, sum)

# check the dependent variable of interest
table(df$good)

prop.table(table(df$good))

```

# Variable Transformation Part 1: Interaction Term

```{r}
mod.fit.Ha = glm(formula = good ~ distance + wind + distance:wind, family = binomial(link = logit), data = df)
summary(mod.fit.Ha)

```

# Likelihood-Ratio Test 1--- using "a"nova
anova($H_0$, $H_1$, test = "Chisq")

```{r}
mod.fit.Ho = glm(formula = good ~ distance + wind, family = binomial(link = logit), data = df)
anova(mod.fit.Ho, mod.fit.Ha, test = "Chisq")

```

p-value 0.02379. this is statistically significant thus we reject null hypothesis.

# Likelihood-Ratio Test 2--- using "A"nova
Anova($H_1$, test = "LR")
```{r}
library(car)
Anova(mod.fit.Ha, test = "LR") # full model including the interaction

```


```{r}
par(mfrow = c(1, 2))
beta0 = mod.fit.Ha$coefficients[1]
beta1 = mod.fit.Ha$coefficients[2]
beta2 = mod.fit.Ha$coefficients[3]
beta1_2 = mod.fit.Ha$coefficients[4]
  

curve(expr = exp(beta0 + beta1*x + beta2*1 + beta1_2*x*1) / (1 + exp(beta0 + beta1*x + beta2*1 + beta1_2*x*1)), xlim = c(20, 70)  )
curve(expr = exp(beta0 + beta1*x + beta2*0 + beta1_2*x*0) / (1 + exp(beta0 + beta1*x + beta2*0 + beta1_2*x*0)), xlim = c(20, 70), add = TRUE )

beta0 = mod.fit.Ho$coefficients[1]
beta1 = mod.fit.Ho$coefficients[2]
beta2 = mod.fit.Ho$coefficients[3]

curve(expr = exp(beta0 + beta1*x + beta2*1) / (1 + exp(beta0 + beta1*x + beta2*1)), xlim = c(20, 70) )
curve(expr = exp(beta0 + beta1*x + beta2*0) / (1 + exp(beta0 + beta1*x + beta2*0)), xlim = c(20, 70), add = TRUE )


```

# Odds Ratio

```{r}
beta.hat = mod.fit.Ha$coefficients[2:4]
c = 1
distance = seq(from = 20, to = 60, by = 10)

OR.wind = exp(c * (beta.hat[2] + beta.hat[3]*distance))
cov.mat = vcov(mod.fit.Ha)[2:4, 2:4]

# Var(beta^_2 + distance*beta^_3)
var.log.OR = cov.mat[2,2] + distance^2*cov.mat[3,3] + 2*distance*cov.mat[2,3]

ci.log.OR.low = c*(beta.hat[2] + beta.hat[3]*distance) - c*qnorm(p=0.975)*sqrt(var.log.OR)
ci.log.OR.up = c*(beta.hat[2] + beta.hat[3]*distance) + c*qnorm(p=0.975)*sqrt(var.log.OR)

round(data.frame(distance = distance, OR.hat = 1/OR.wind, OR.low = 1/exp(ci.log.OR.up),
                 OR.up = 1/exp(ci.log.OR.low)), 2)

```

# Quadratic Term:

```{r}
model_ha = glm(formula = good ~ distance + I(distance^2), family = binomial(link = logit),
    data = df)

model_ho = glm(formula = good ~ distance, family = binomial(link = logit),
    data = df)

summary(model_ha)
```

# Likelihood Ratio Test --- anova, Anova

```{r}

anova(model_ho, model_ha, test = "Chisq")

library(package = car)
Anova(model_ha)


```

# Plot the Estimated Model

```{r}
# refer p.74
w = aggregate(formula = good ~ distance, data = df, FUN = sum)
n = aggregate(formula = good ~ distance, data = df, FUN = length)
w.n = data.frame(distance = w$distance, success = w$good, trials = n$good, 
                 proportion = round(w$good/n$good, 4))

beta.hat.ha  = model_ha$coefficients
beta.hat.ho  = model_ho$coefficients

symbols(x = w$distance, y = w$good/n$good, circles = sqrt(n$good), inches = 0.5,
        xlab = "Distance (yards)", ylab = "Estimated probability", xlim = c(10, 70),
        panel.first = grid(col = "gray", lty = "dotted"))

curve(expr = exp(beta.hat.ha[1] + beta.hat.ha[2]*x + beta.hat.ha[3]*x^2) / (1 + exp(beta.hat.ha[1] + beta.hat.ha[2]*x + beta.hat.ha[3]*x^2)), xlim = c(10, 70), col = "red", add = TRUE ) # full model
curve(expr = exp(beta.hat.ho[1] + beta.hat.ho[2]*x) / (1 + exp(beta.hat.ho[1] + beta.hat.ho[2]*x)), xlim = c(10, 70), col = "blue", add = TRUE ) # null model

```

# Categorical Explanatory Variables

```{r}

tomato = read.table(file = "TomatoVirus.csv", header = TRUE, sep = ",")
head(tomato)

class(tomato$Control)

levels(tomato$Control)

tomato$Infest = factor(tomato$Infest)

```


```{r}
mod.fit = glm(formula = Virus8/Plants ~ Infest + Control, family = binomial(link = logit),
              data = tomato, weights = Plants)
summary(mod.fit)

```

base case is control "B".
It's good to write down the model.
























