---
title: "two sample t-test"
author: "K Iwasaki"
date: "September 11, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}

S = read.csv("united_states_senate_2014_v2.csv")
summary(S)

```

# Is there a difference between the amount of money a senator raises and the amount spent?
**Dependent t-test (paird t-test)**

0. Analyse the problem.

There is a claer pairing in this case. Money raised vs. money spent for each senator.
Since we have a large sample, we can use CLT to justify parametric test.

1. Construct hypothesis.

$H_0$ is there is no difference between them.

2. Check the assumption of normality

The distributions for the money raised and spent are not normal but should be fine because of CLT.
```{r}
hist(S$Campaign.Money.Raised..millions.of..., breaks = 20)

hist(S$Campaign.Money.Spent..millions.of..., breaks = 20)

diff = S$Campaign.Money.Raised..millions.of... - S$Campaign.Money.Spent..millions.of...

hist(diff, breaks = 20)

```

3. Check the assumption of equal variances.

Since they are paird. This is ok. There is onlly one distribution to test.

4. Run the t.test
```{r}
t.test(S$Campaign.Money.Raised..millions.of..., S$Campaign.Money.Spent..millions.of..., paired = T)

```

p-value is very small. Reject null hypothesis that there is no difference in the money raised and money spent for the senatros.

5. Calculate Effect Size

The differences in the mean is 1.418

# Do female Democratic senators raise more or less money than female Republican senators?
**Independent t-test **

0. Analyze the problem

The data we are interested in are not paired.

1. Construct hypothesis.

$H_0$ is there is no difference between average money raised by female democratic senators and average money raised by female republican senators. Two tailed test since we don't know less or more.
Also, we doon't have many observation.

2. Check the assumpiton of normality

```{r}
female_democrat_mask = (S$Gender == "Female") & (S$Party == "Democrat")
female_republican_mask = (S$Gender == "Female") & (S$Party == "Republican")

female_democ_money = S[female_democrat_mask, ]$Campaign.Money.Raised..millions.of...
female_repub_money = S[female_republican_mask, ]$Campaign.Money.Raised..millions.of...

hist(female_democ_money, breaks = 20)
hist(female_repub_money, breaks = 20)

```
3. Check the assumption of equal variances

```{r}
library(car)
leveneTest(Campaign.Money.Raised..millions.of... ~ Party, data = S[S$Gender=="Female",])

```

```{r}
wilcox.test(Campaign.Money.Raised..millions.of... ~ Party, data = S[S$Gender=="Female",])
```


```{r}
t.test(Campaign.Money.Raised..millions.of... ~ Party, data = S[S$Gender=="Female",])
```






