---
title: "hw08"
author: "K Iwasaki"
date: "June 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/Users/K/Desktop/Berkeley/00_Academics/01_2017 Summer/W203 STATS/hw08")
#closeAllConnections()
#rm(list=ls())
```

### a. Examine the skipped variable and argue whether or not a t-test is valid for this scenario
As shown in the chart below, the skipped variable is skewed to the right. Thus, t-test is not valid for this scenario

```{r}
load("GPA1.RData")
head(data)
hist(data$skipped)
table(data$skipped)

```

### b. How would your answer to part a change if Mr. Lemmon selected dormitory rooms at random, then interviewed all occupants in the rooms he selected?

As long as Mr. Lemmon selects sufficiently large samples (n > 30), t-test is valid for this scenario

### c. Provide an argument for why you should choose a 2-tailed test in this instance, even if you are hoping to demonstrate that MSU students skip more than 1 lecture per week.

Because we don't know whether the MSU students skip more or less than 1 lecture per week on the average, we should carry out a 2-tailed test in this instance in order to make the test more conservative.

## d. Conduct the t-test using the t.test function and interpret every component of the results.
```{r}
t.test(data$skipped, alternative = "two.sided", mu = 1, conf.level = 0.95)

```
- data$skipped is used for t-test
- t-test is 0.83142, degree of freedom is 140 which is the number of sample 141 - 1, p-value for two-tailed test is 0.4072
- what the p-value means is that if null hypothesis, the mean is larger than 1, is true, the chance of getting the sample mean and sample sd will be 41% purely as a result of sampling error, which is quite high.
- since p-value is higher than significance level 0.05, we don't reject null hypothesis.
- 95% confidence interval is also calculated as upperlimit 1.257 and lowerlimit 0.895
- lastly, the mean of sample is calculated: 1.0762

### e. Show how you would compute the t-statistic and p-value manually (without using t.test), using the pt function in R.

```{r}
# calculate statistics
s.mean <- mean(data$skipped)
s.stdev <- sd(data$skipped)
n <- length(data$skipped)

# critical value
a <- 1 - 0.95
(c.value <- abs(qt(p = a/2, df = n-1)))

# t-stat
(t <- (s.mean - 1) / (s.stdev / sqrt(n) ))

# p-value
2 * pt(q= -t, df = n-1)

# standard error
(s.error <- s.stdev / sqrt(n))

# 95% confidence interval
c(s.mean - c.value * s.error, s.mean + c.value * s.error)

```


### f. Construct a 99% confidence interval for the mean number classes skipped by MSU students in a week.
```{r}

# critical value 99%
a <- 1 - 0.99
(c.value <- abs(qt(p = a/2, df = n-1)))

# standard error
(s.error <- s.stdev / sqrt(n))

# 99% confidence interval
c(s.mean - c.value * s.error, s.mean + c.value * s.error)

```


### g. Can you say that there is a 99% chance the population mean falls inside your confidence interval?

No. Confidence is on methodology. 99% confidence interval means that if we use the same methodology to construct 99% confidence intervals. 99% of the intervals capture the population mean.