---
title: "hw12"
author: "K Iwasaki"
date: "August 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#closeAllConnections()
#rm(list=ls())
setwd("C:/Users/K/Desktop/Berkeley/00_Academics/W203 STATS/hw12")

library(lmtest)
library(sandwich)
library(stargazer)
library(car)

```

# 0. Quick data exploration and data clean up

```{r}
data = read.table("videos.txt", fill = TRUE, header = TRUE)
head(data)
nrow(data)

# check number of na in each col
sapply(data, function(x) sum(is.na(x))) 

# transform factor into numeric
data$views = as.numeric(levels(data$views))[data$views]
data$length = as.numeric(levels(data$length))[data$length]

# check number of na in each col again
sapply(data, function(x) sum(is.na(x))) 

# remove rows containing NA in the three columns
data = data[complete.cases(data[, c("length", "views", "rate")]),]
nrow(data)

scatterplotMatrix(data[, c("views", "length", "rate")])

# examine views column
summary(data$views)
hist(data$views, breaks = 1000)
# extremely skewed to the right

# check outliers
data[data$views > 500000, ]

# examine length column
summary(data$length)
hist(data$length, breaks = 1000)
# extremely skewed to the right

# check outliers
data[data$length > 1000, ]

# examine rate
summary(data$rate)
hist(data$rate, breaks = 1000)
```


# 1. Fit a linear model predicting the number of views (views), from the length of a video (length) and its average user rating (rate).


```{r}
(model1 = lm(views ~ length + rate, data = data))
```

```{r}
plot(model1)
```

# 2. Using diagnostic plots, background knowledge, and statistical tests, assess all 6 assumptions of the CLM. When an assumption is violated, state what response you will take.

1. Linear population model. This assumption is ok.
2. Random sampling. There is not much information about how this data is collected.I assume this assumption isn't violated.
3. No perfect multicolliearity. The plot below doesn't show perfect multicolliearity.

```{r}
plot(data$length, data$rate)
```
4. Zero-conditional mean. Look at the Residuals vs. Fitted plot. Smooth line in the chart is around 0. It seems this assumption is not violated.

5. Homoskedasticity. Look at the Residuals vs. Fitted plot and Scale-location plot. A bit difficult to say if the model shows homoskedasticity or not. To double-check, run a Breaush-Pagan test. p-value is large so we don't reject the null hypothesis that the model has homoskedasticity. Thus this assumption is not violated.

```{r}
bptest(model1)
```


6. Normality of error term. Look at the q-q plot. This assumption is violated. Since the sample size is 6292, we assume normality of our sampling distributions from the large sample size. Thus, no action on this violation.


# 3. Generate a printout of your model coefficients, complete with standard errors that are valid given your diagnostics. Comment on both the practical and statistical significance of your coefficients.

Use robust standard errors since it's good practice and I wasn't sure about homoskedasticity.
```{r}
coeftest(model1, vcov = vcovHC)
```

Look at the predictor variables.

## rate
p-value is quite small and thus statistically significant. coefficient is 2152. What this means is that holding length is fixed, another 1 point on rate are associated with 2152 views. This is pratically significant.

## length
p-value is quite large as 0.265 and thus this predictor variable is not statistically significant. Coefficient is only 1.9 and not pratically significant.



























