---
title: "Homework Exercise 1"
author: "K Iwasaki"
date: "May 18, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

W203 Statistics for Data Science
========================================================
Unit 1 Homework
========================================================


## Exercise

Load the dataset found in the file, cars.csv.

```{r load}
setwd("C:/Users/K/Desktop/Berkeley/00_Academics/01_2017 Summer/W203 STATS/hw01")
data <- read.csv('cars.csv')

# check the data
head(data)

nrow(data)

?cars


```


1. What are the variables in the file?
```{r}
str(data)
```


2. Find the mean, median, minimum, maximum, 1st quartile and 3rd quartile for the mpg variable.
```{r}
summary(data$mpg)
```


3. Create a histogram of the mpg variable. 
```{r}

hist(data$mpg, breaks = 10:35 - 0.5, 
     main = 'Distribution of Variable "mpg"', xlab = NULL)

```


4. What is the standard deviation of mpg variable?
```{r}
std = sd(data$mpg)
std
```


5. What is the variance of mpg variable?
```{r}
variance = var(data$mpg)
variance
```

6. What is the relationship of the standard deviation to the variance? Why does the standard deviation and variance of the mpg variable differ? 
```{r}
# create a table and compare the values visually
mpg_df <- data.frame(variable=c("standard deviation", "variance"), val = c(std, variance))

barplot(mpg_df$val, names.arg = mpg_df$variable)

# The standard deviation is the square root of the variance by definition.
# Let's check if this is true here.

std_power2 <- std **2
check <- std_power2 == variance
check

  
# viualize this
new.row1 <- data.frame(variable = "std_power2", val = std_power2)
mpg_df <- rbind(mpg_df, new.row1)

barplot(mpg_df$val, names.arg = mpg_df$variable)


```


7. How many data points are there for the cyl variable? 
```{r}
length(data$cyl)
summary(data$cyl) # note there is two NAs

```

8. What is the mean of the cyl variable?
```{r}
mean(data$cyl, na.rm=T)

```
